Covariate analysis of CMC RNASeq data reprocessed with Star-Feature counts pipeline (ACC and DLPFC with TMM normlaisation and SVA)

### 2016-12-06 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM.Rmd
Added new desing for SVA based adjustments

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gc_content"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)
GC.CONT = Ensemble2HGNC %>%
  dplyr::select(ensembl_gene_id, percentage_gc_content) %>%
  unique
rownames(GC.CONT) = GC.CONT$ensembl_gene_id
```

@@

```{r dataExplore, fig.width = 5, fig.height = 5}
my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1))

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2)) 
p <- p + geom_point(aes(color=Institution, shape = Dx.Tissue, size=RIN)) 
p <- p + theme_bw() + theme(legend.position="top")
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=1)
p
```

>>>>

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gc_content", "transcript_length"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)
GC.CONT = Ensemble2HGNC %>%
  dplyr::select(ensembl_gene_id, percentage_gc_content) %>%
  unique
rownames(GC.CONT) = GC.CONT$ensembl_gene_id

Gx.LEN = Ensemble2HGNC %>%
  dplyr::select(ensembl_gene_id, transcript_length) %>%
  group_by(ensembl_gene_id) %>%
  summarise(transcript_length = max(transcript_length, na.rm=T)) %>%
  as.data.frame()
rownames(Gx.LEN) = Gx.LEN$ensembl_gene_id
```

@@

```{r dataExplore, fig.width = 5, fig.height = 5}
my.theme <- theme(legend.position = 'top', axis.text.x = element_text(angle = 90, hjust = 1),
                  plot.title=element_text(hjust=0.5))

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2)) 
p <- p + geom_point(aes(color=Institution, shape = Dx.Tissue, size=RIN)) 
p <- p + theme_bw() + theme(legend.position = "top")
# p <- p + geom_text(aes(label= SampleID), size=4, hjust=1)
p
```

### 2016-12-06 code/Rmd/CMC_ACC_And_DLPFC_StarAligned_TMM_SVA.Rmd
Added new desing for SVA based adjustments

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download reprocessed counts (ACC)
COUNT_ACC_ID = 'syn7217917';
ALL_USED_IDs = COUNT_ACC_ID
COUNT_ACC = read.table(synGet(COUNT_ACC_ID)@filePath, header=T, sep='\t', check.names = F, nrows = 56632) %>%
  dplyr::select(-chr, -start, -end, -strand, -length)

@@

## Metadata with specific information for RNASeq processing
metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", 
                          "Dx", "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?", "ACC_RNA_report: Library Batch",
                          "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?", "DLPFC_RNA_report: Library Batch",
                          'DNA_genotyping_report: Genotyping Sample_ID')

@@

METADATA_ACC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, ACC_RNA_isolation__Exclude_,
                ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN, ACC_RNA_report__Exclude_,
                ACC_RNA_report__Library_Batch, DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = ACC_RNA_isolation__Exclude_, 
                SampleID = ACC_RNA_isolation__Sample_RNA_ID, 
                RIN = ACC_RNA_isolation__RIN, 
                ReportExclude = ACC_RNA_report__Exclude_, 
                LibraryBatch = ACC_RNA_report__Library_Batch,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'ACC')

METADATA_DLPFC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, DLPFC_RNA_isolation__Exclude_,
                DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN, DLPFC_RNA_report__Exclude_,
                DLPFC_RNA_report__Library_Batch, DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
                RIN = DLPFC_RNA_isolation__RIN, 
                ReportExclude = DLPFC_RNA_report__Exclude_, 
                LibraryBatch = DLPFC_RNA_report__Library_Batch,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'DLPFC')

METADATA = rbindlist(list(METADATA_ACC, METADATA_DLPFC), use.names = T, fill = T) %>%
  droplevels() %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch)) %>%
  left_join(ANCESTRY)

@@

@@

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')
```

@@

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```

### Normalisation (custom design)
Normalising COUNTS with a design matrix obtained from an iterative normalisation analysis (see syn7503251 for more details).

Here Institution, Gender, RIN, Age\\_of\\_Death, PMI, EV.4 were used as fixed effects conditioned on Dx.Tissue with Individual\\_ID as random effects.
```{r customNorm, results='asis'}
# Primary variable of interest
postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'RIN', 'Age_of_Death', 'PMI', 'EV.4');

@@

# Find PC of residual gene expression and significant covariates that are highly correlated with PCs
residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION, 
                                              COVARIATES[, ind.cov, drop=F], 
                                              'adjusted design(voom-normalized)',
                                              isKeyPlot=TRUE)
```
### Residual calculation
Calculate weighted residuals and add back "Dx.Tissue" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

>>>>

# Github link
thisRepo <- getRepo(repository = "th1vairam/Brain_Reg_Net", ref="branch", refName='CMC')
thisFile <- getPermlink(repository = thisRepo, repositoryPath=paste0('code/Rmd/',thisFileName))
```
### Data download
#### Obtain count matrix and metadata from synapse.
### Data download
#### Obtain count matrix and metadata from synapse.
```{r download.data, cache=TRUE}
# Download reprocessed counts (ACC)
COUNT_ACC_ID = 'syn7217917';
ALL_USED_IDs = COUNT_ACC_ID
COUNT_ACC = read.table(synGet(COUNT_ACC_ID)@filePath, header=T, sep='\t', check.names = F, nrows = 56632) %>%
  dplyr::select(-chr, -start, -end, -strand, -length)

@@

## Metadata with specific information for RNASeq processing
metadataVarOfInterest = c("Individual ID", "Institution", "Gender", "Age of Death", "PMI (in hours)", "Dx",

                          "ACC_RNA_isolation: Exclude?", "ACC_RNA_isolation: Sample RNA ID",
                          "ACC_RNA_isolation: RIN", "ACC_RNA_report: Exclude?", "ACC_RNA_report: Library Batch",
                          "ACC_RNA_report: Ribozero Batch", "ACC_RNA_report: Flowcell Batch", 
                          "ACC_RNA_report: Mapped Reads", "ACC_RNA_report: Intragenic Rate",
                          "ACC_RNA_report: Intronic Rate", "ACC_RNA_report: Intergenic Rate",
                          "ACC_RNA_report: Genes Detected", "ACC_RNA_report: Transcripts Detected",
                          "ACC_RNA_report: Expression Profiling Efficiency", "ACC_RNA_report: rRNA Rate",
                          "ACC_RNA_report: Total Reads", "ACC_RNA_report: Percent Aligned",

                          "DLPFC_RNA_isolation: Exclude?", "DLPFC_RNA_isolation: Sample RNA ID",
                          "DLPFC_RNA_isolation: RIN", "DLPFC_RNA_report: Exclude?", "DLPFC_RNA_report: Library Batch",
                          "DLPFC_RNA_report: Ribozero Batch", "DLPFC_RNA_report: Flowcell Batch",
                          "DLPFC_RNA_report: Mapped Reads", "DLPFC_RNA_report: Intragenic Rate",
                          "DLPFC_RNA_report: Intronic Rate", "DLPFC_RNA_report: Intergenic Rate", 
                          "DLPFC_RNA_report: Genes Detected", "DLPFC_RNA_report: Transcripts Detected", 
                          "DLPFC_RNA_report: Expression Profiling Efficiency", "DLPFC_RNA_report: rRNA Rate",
                          "DLPFC_RNA_report: Total Reads", "DLPFC_RNA_report: Percent Aligned",

                          "DNA_genotyping_report: Genotyping Sample_ID",
                          "DNA_genotyping_report: Exclude?")

@@

METADATA_ACC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
                ACC_RNA_isolation__Exclude_, ACC_RNA_isolation__Sample_RNA_ID, ACC_RNA_isolation__RIN,
                ACC_RNA_report__Exclude_, ACC_RNA_report__Ribozero_Batch, ACC_RNA_report__Library_Batch, 
                ACC_RNA_report__Flowcell_Batch, ACC_RNA_report__Mapped_Reads, ACC_RNA_report__Intragenic_Rate,
                ACC_RNA_report__Intronic_Rate, ACC_RNA_report__Intergenic_Rate, ACC_RNA_report__Genes_Detected,
                ACC_RNA_report__Transcripts_Detected, ACC_RNA_report__Expression_Profiling_Efficiency, 
                ACC_RNA_report__rRNA_Rate, ACC_RNA_report__Total_Reads, ACC_RNA_report__Percent_Aligned,
                DNA_genotyping_report__Genotyping_Sample_ID, DNA_genotyping_report__Exclude_) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = ACC_RNA_isolation__Exclude_, 
                SampleID = ACC_RNA_isolation__Sample_RNA_ID, 
                RIN = ACC_RNA_isolation__RIN, 
                ReportExclude = ACC_RNA_report__Exclude_, 
                RibozeroBatch = ACC_RNA_report__Ribozero_Batch,
                LibraryBatch = ACC_RNA_report__Library_Batch,
                FlowcellBatch = ACC_RNA_report__Flowcell_Batch, 
                MappedReads = ACC_RNA_report__Mapped_Reads, 
                IntragenicRate = ACC_RNA_report__Intragenic_Rate,
                IntronicRate = ACC_RNA_report__Intronic_Rate, 
                IntergenicRate = ACC_RNA_report__Intergenic_Rate, 
                GenesDetected = ACC_RNA_report__Genes_Detected,
                TranscriptsDetected = ACC_RNA_report__Transcripts_Detected, 
                ExpProfEfficiency = ACC_RNA_report__Expression_Profiling_Efficiency, 
                rRNARate = ACC_RNA_report__rRNA_Rate, 
                TotalReads = ACC_RNA_report__Total_Reads, 
                PercentAligned = ACC_RNA_report__Percent_Aligned,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID,
                GenotypingExclude = DNA_genotyping_report__Exclude_) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'ACC')

METADATA_DLPFC = METADATA %>%
  dplyr::select(Individual_ID, Institution, Gender, Age_of_Death, PMI__in_hours_, Dx, 
                DLPFC_RNA_isolation__Exclude_, DLPFC_RNA_isolation__Sample_RNA_ID, DLPFC_RNA_isolation__RIN,
                DLPFC_RNA_report__Exclude_, DLPFC_RNA_report__Ribozero_Batch, DLPFC_RNA_report__Library_Batch, 
                DLPFC_RNA_report__Flowcell_Batch, DLPFC_RNA_report__Mapped_Reads, DLPFC_RNA_report__Intragenic_Rate,
                DLPFC_RNA_report__Intronic_Rate, DLPFC_RNA_report__Intergenic_Rate, DLPFC_RNA_report__Genes_Detected,
                DLPFC_RNA_report__Transcripts_Detected, DLPFC_RNA_report__Expression_Profiling_Efficiency, 
                DLPFC_RNA_report__rRNA_Rate, DLPFC_RNA_report__Total_Reads, DLPFC_RNA_report__Percent_Aligned,
                DNA_genotyping_report__Genotyping_Sample_ID, DNA_genotyping_report__Exclude_) %>%
  dplyr::rename(PMI = PMI__in_hours_, 
                IsolationExclude = DLPFC_RNA_isolation__Exclude_, 
                SampleID = DLPFC_RNA_isolation__Sample_RNA_ID, 
                RIN = DLPFC_RNA_isolation__RIN, 
                ReportExclude = DLPFC_RNA_report__Exclude_, 
                RibozeroBatch = DLPFC_RNA_report__Ribozero_Batch,
                LibraryBatch = DLPFC_RNA_report__Library_Batch,
                FlowcellBatch = DLPFC_RNA_report__Flowcell_Batch, 
                MappedReads = DLPFC_RNA_report__Mapped_Reads, 
                IntragenicRate = DLPFC_RNA_report__Intragenic_Rate,
                IntronicRate = DLPFC_RNA_report__Intronic_Rate, 
                IntergenicRate = DLPFC_RNA_report__Intergenic_Rate, 
                GenesDetected = DLPFC_RNA_report__Genes_Detected,
                TranscriptsDetected = DLPFC_RNA_report__Transcripts_Detected, 
                ExpProfEfficiency = DLPFC_RNA_report__Expression_Profiling_Efficiency, 
                rRNARate = DLPFC_RNA_report__rRNA_Rate, 
                TotalReads = DLPFC_RNA_report__Total_Reads, 
                PercentAligned = DLPFC_RNA_report__Percent_Aligned,
                GenotypingSampleID = DNA_genotyping_report__Genotyping_Sample_ID,
                GenotypingExclude = DNA_genotyping_report__Exclude_) %>%
  dplyr::filter(SampleID %in% colnames(COUNT)) %>%
  dplyr::mutate(Tissue = 'DLPFC')

METADATA = rbindlist(list(METADATA_ACC, METADATA_DLPFC), use.names = T, fill = T) %>%
  droplevels() %>%
  dplyr::mutate(LibraryBatch = as.factor(LibraryBatch)) %>%
  left_join(ANCESTRY) %>%
  unique

@@

writeLines('Following samples are marked excluded in genotyping')
writeLines(paste(METADATA$SampleID[METADATA$GenotypingExclude == 1], collapse = ', '))
METADATA <- METADATA  %>% filter(GenotypingExclude == 0)

@@

# Add in RIN^2 values
COVARIATES$RIN2 = COVARIATES$RIN^2
ContCovariates = c(ContCovariates, 'RIN2')

## Get GC content from biomart
backgroundGenes = data.frame(ensembl_gene_id = rownames(COUNT))

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol", "percentage_gc_content", "transcript_length"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)
GC.CONT = Ensemble2HGNC %>%
  dplyr::select(ensembl_gene_id, percentage_gc_content) %>%
  unique
rownames(GC.CONT) = GC.CONT$ensembl_gene_id

Gx.LEN = Ensemble2HGNC %>%
  dplyr::select(ensembl_gene_id, transcript_length) %>%
  group_by(ensembl_gene_id) %>%
  summarise(transcript_length = max(transcript_length, na.rm=T)) %>%
  as.data.frame()
rownames(Gx.LEN) = Gx.LEN$ensembl_gene_id
```

@@

Significant covariates to adjust at FDR 0.1 are `r preAdjustedSigCovars$significantCovars`
```{r preAdjustedSigCovars.NULL, fig.width=20, fig.height=12}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (custom design)
Normalising COUNTS with a design matrix obtained from an iterative normalisation analysis (see syn7503251 for more details).

Here Dx.Tissue, Institution, Gender, Intronic Rate, Intergenic Rate, Transcripts Detected, Mapped Reads, Percent Aligned, RIN, PMI, Age\\_of\\_Death, rRNA Rate, Ribozero Batch and EV.1, EV.2, EV.3, EV.4 and EV.5 were chosen as fixed effects and Individual\\_ID as random effects.
```{r customNorm, results='asis'}
# Primary variable of interest
postAdjustCovars = c('Dx.Tissue', 'Institution', 'Gender', 'IntronicRate', 'IntergenicRate', 'TranscriptsDetected', 
                     'MappedReads', 'PercentAligned', 'RIN', 'PMI', 'Age_of_Death', 'rRNARate', 'RibozeroBatch',
                     'EV.1', 'EV.2', 'EV.3', 'EV.4', 'EV.5');

@@

### Residual calculation
Calculate weighted residuals and add back "Dx.Tissue" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("Dx.Tissue", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```
Coexpression of genes 
```{r coexp2, cache=FALSE, fig.height=5, fig.width=5}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```
